---
title: "Exploratory Analysis of COVID 19"
date: '2022-08-04'
output:
  html_document: default
  pdf_document: default
  word_document: default
---
## Libraries Used
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Problem

With COVID 19 case and deaths numbers on the rise, we would like to evaluate which state is being hit the hardest.

## Data Description

The data used for the analysis came from Johns Hopkins data repository on GitHub

## Import Data

```{r get_data}
url_in<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names<- c("time_series_covid19_confirmed_US.csv", 
               "time_series_covid19_confirmed_global.csv", 
               "time_series_covid19_deaths_US.csv", 
               "time_series_covid19_deaths_global.csv", 
               "time_series_covid19_recovered_global.csv")
urls<-str_c(url_in, file_names)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

``` {r import_data}
global_cases<-read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read.csv(urls[1])
US_deaths <- read.csv(urls[3])
```

## Tidy Data

```{r tidy_data_global_data}
global_cases <- global_cases%>%
  select(-c(Lat, Long))%>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region'), 
               names_to = "date",
               values_to = "cases")

global_deaths <- global_deaths%>%
  select(-c(Lat, Long))%>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region'), 
               names_to = "date",
               values_to = "deaths")

global_cases
global_deaths

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global
```
```{r tidy_data_US_data}
US_cases <- US_cases %>%
  pivot_longer(cols = -c('UID':'Combined_Key'),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(gsub("X", "", date))) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(gsub("X", "", date))) %>%
  select(-c(Lat, Long_))
US_cases
US_deaths

US<- US_cases %>%
  full_join(US_deaths)
```
``` {R tidy_all}
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
global <- global%>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, 
         cases, deaths, Population, 
         Combined_Key)

```
### Discussion for missing data


## Analysis

```{r analysis}

US_by_state <- US %>%
  group_by (Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state %>%
  group_by (Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

#Visualization number 1
US_totals %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US", y = NULL)

#visualisation 2
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title = str_c("COVID19 in ", state), y = NULL)

US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases) ,
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))

US_totals %>%
  ggplot(aes(x = date, y = new_cases))+
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = new_deaths, color = "new_deaths"))+
  geom_point(aes(y = new_deaths, color = "new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US", y = NULL)

US_state_totals <- US_by_state %>%
  group_by(Province_State)%>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000* cases/population,
            deaths_per_thou = 1000* deaths/population) %>%
  filter(cases>0, population >0)

US_state_totals %>%
  slice_min(deaths_per_thou, n = 10)%>%
  select (deaths_per_thou, cases_per_thou, everything())

US_state_totals %>%
  slice_max(deaths_per_thou, n = 10)%>%
  select (deaths_per_thou, cases_per_thou, everything())
```


```{r model}

mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

summary(mod)

US_state_totals %>% slice_min(cases_per_thou)
US_state_totals %>% slice_max(cases_per_thou)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred %>% 
  ggplot()+
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue")+
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```
```{r additional analysis}

state <- "Ohio"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = new_deaths, color = "new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title = str_c("COVID19 in ", state), y = NULL)

state <- "Colorado"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y = cases))+
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = new_deaths, color = "new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title = str_c("COVID19 in ", state), y = NULL)

```


Reviewing the model created, the p value is below .005 supporting that the model is statistically significant. When the Residuals vs Fitted plot is looked at, we can conclude that a linear regression model is appropriate for the dataset since the redline does not deviate from the horizontal dotted line substantially. Figure 7 helps visualize the model that was created and supports that there is a linear relationship between incident rate and murder rate, as the amount of shootings goes up so does the amount of murders in Brooklyn.

## Bias

When reviewing the data set, we wanted to form a question that specifically did not require race to be evaluated in order for it to be answered. As it relates to gun violence, we did not believe to add significant value to an analysis because it has been shown that race does not necessarily correlate to violence and to include it in an analysis would only serve to introduce complexity. To state it a different way, there are concentrations of certain races, for various other factors, in these areas and as a result there will be a higher amount of those races committing these crimes. To mitigate this, race was removed from the data subset that would be analyzed. 

## Discussion and Summary

The analysis has successfully identified the highest concentration of gun crime in NYC and provided visualizations of when shootings occur by hour, where the occur most often and who, by sex and age, is committing the crimes. Additionally, the analysis provides a linear model to support that by reducing all gun violence, there would be a decrease in violent crime (murders). 

The analysis should be used to develop actionable items that can be taken to reduce gun violence. These actions might include, but not be limited to, increasing police presence during hours where gun violence is more frequent and increasing funding to inner city programs that are focusing efforts around youth and their maturity from teenage years through young adult hood. 

# Future Analysis

Further steps should be taken to conduct the same analysis on the other boroughs in NYC. While it is expected the similar conclusions would be drawn, it should not be stated as fact without the analysis being completed. Additionally, this analysis should be completed on an annual basis as mitigation to the violence are rolled out to evaluate effectiveness of the efforts. Finally, the analysis should be continually updated and expanded as more data is included to ensure the right steps are being taken to curb the present issue of gun violence. 
